---
title: 'Reproducible Research: Peer Assessment 1'
author: "Ricardo Leon"
date: "Sun Jan 12 2020"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
source("utils.R")
data <- load_data()
```


## What is mean total number of steps taken per day?
```{r}
data_per_day <- get_total_steps_per_day(data)

ggplot(data_per_day, aes(x = total_steps)) +
    geom_histogram(bins = 50) +
    ggtitle("Frequency of total steps") +
    xlab("Total Steps") +
    ylab("Frequency")
overall_mean <- mean(data_per_day$total_steps, na.rm = TRUE)
overall_median <- median(data_per_day$total_steps, na.rm = TRUE)
```

- The mean value for the total steps per day is `r overall_mean` and the median `r overall_median`.


## What is the average daily activity pattern?
```{r}
data_per_interval <- get_mean_per_interval(data)

ggplot(data_per_interval, aes(x=interval, y=mean_value)) +
    geom_line() +
    ggtitle("Average number of steps by interval") +
    xlab("Interval") +
    ylab("Number of steps")
max_interval <- (data_per_interval %>%
                    filter(mean_value == max(mean_value)))$interval
```

- The maximum number of steps is in interval `r max_interval`.


## Imputing missing values
```{r}
data_no_na <- impute_data_with_median(data)
data_per_day_no_na <- get_total_steps_per_day(data_no_na)

total_missing_values <- sum(is.na(data$steps))
total_zero_values <- sum(data$steps==0, na.rm = TRUE)
ggplot(data_per_day_no_na, aes(x = total_steps)) +
        geom_histogram(bins = 50) +
        ggtitle("Frequency of total steps") +
        xlab("Total Steps") +
        ylab("Frequency")
overall_mean_no_na <- mean(data_per_day_no_na$total_steps)
overall_median_no_na <- median(data_per_day_no_na$total_steps)
```

- There are `r total_missing_values` missing values.
- Missing values were replaced by the median of each day.
- With imputed values, the mean value for the total steps per day is `r overall_mean_no_na` and the median `r overall_median_no_na`.
- Imputing missing values with the value zero (median individual value) makes both, the mean and the median of the total number of steps, decrease because the number of observations increases but no value was added.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
data_per_interval_and_type <- mean_by_day_type(data_no_na)

ggplot(data_per_interval_and_type, aes(x = interval, y = mean_value)) +
        geom_line() +
        facet_grid(type ~ .) +
        ggtitle("Plot steps-interval by day type") +
        xlab("Interval") +
        ylab("Number of steps")
```